Data Migration under a Changing Schema

### 2023-08-06 2022Migration/articleMigrationFACS.tex
Simplifying the explanation

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
   while transferring data in a controlled fashion.
   Typically, a migration strategy (e.g. case by case, customer by customer, or in batches) is in place
   to ensure the preservation of valuable data and the quality of the migration.
   To fully automate every incremental data migration is unrealistic.
   Smaller increments stand a higher chance of being automated, but there will always be cases where some human intervention is required.
   Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.
   Consequently, a migration engineer must have the freedom to tailor any automatic approach to migration to suit the particulars.

   Complications that require manual intervention arise mostly as a result of data pollution.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that cannot be captured in constraints on the data set,
   such as a street address that has become obsolete because someone has moved without notification.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category is about violations of constraints from the schema.
   Such violations do not occur in our theory because we assume the existing system was generated by Ampersand,
   so the data set satisfies all constraints in the schema.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the desired system.
   In that case, a migration engineer must ensure that the desired system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The last category of data pollution, erroneous constraints, must be fixed by removing them or replacing them by the correct constraints in the desired system.

   This illustrates that automating data migrations may require user intervention,
   either by a migration engineer or by end-users.
   In our approach, such user interventions are given some time by
   defining two distinct moments:
\begin{enumerate}
   \item the moment of transition (the MoT), i.e. the moment that changed functionality is made available to users;
   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.
   This leaves the business with the intended system: a successful migration.
\end{enumerate}

   During the time between the MoT and MoC,
   the existing system and the desired system are kept alive, side by side, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}

   The migration requires a third schema, which specifies the migration itself.
   The migration schema comprises the schemas of both the existing system and the desired system.
   The rules of the migration schema will cause the right data to be copied correctly into the desired system.
   This migration schema preserves data from the existing system, it may introduce new data automatically (generated from the existing data set),
   and it may require users to introduce new data (after the MoT and before the MoC).
   So the migration schema is a (disjoint) union of the existing and the desired schemas,
   which is presented to the migration engineer in the form of source code,
   so the migration engineer can change everything she wants to suit the particulars of the data migration.

%During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
%This allows the business to finish transactions in the existing system while the desired system is up and running.
%Similarly, the existing system must execute transactions from the desired system, just in case the business calls off the migration.
%Rules in the migration system that require user interaction are given time by requiring that the migration phase ends
%only after all migration rules are at rest.

@@

   \begin{definition}[Data set]
   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance \subseteq  {\Pair{\Atoms}{\Concepts}}$ that satisfies:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\instance A\ \wedge\ b\instance B
   \label{eqn:wellTypedEdge}
\end{eqnarray}
\end{definition}
   Looking at the example,
   equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that the Person Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
   %To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.

@@

   For a different kind of constraint, the system is told what to do in order to fix it.
   Such constraints are called enforcement rules.
   We use just one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
   Its syntax is as follows:
\begin{align}
      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
\end{align}

@@

   Let us look what happens at runtime.
   Initially, when {\tt copied\_isPartOf} is still empty, the difference between {\tt old.isPartOf} and {\tt copied\_isPartOf} equals {\tt old.isPartOf}.
   So, {\tt new.isPartOf} will contain the pairs of {\tt old.isPartOf} by rule~\ref*{difference}.
   Then, the engine inserts the same pairs into {\tt copied\_isPartOf} because of rule~\ref*{fill copies with new.isPartOf}.
   After this, triples with the relation {\tt new.isPartOf} can be inserted and deleted freely.

>>>>

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}
   To automate the migration as much as possible,
   we also deploy a migration system,
   which has its own schema to specify the migration itself.
   The migration system comprises the schemas of both the existing system and the desired system,
   to implement the copying of data.
   It also implements the migration strategy,
   automated policies to resolve data pollution,
   and other rules at the discretion of the migration engineer.
   To fully automate every data migration is unrealistic.
   A migration engineer can minimize manual work by minimizing the size of the increments,
   but she cannot eliminate it completely.
   Data pollution, new business rules, or known issues in the existing system
   may occasionally require the migration engineer to tailor the migration script to specific needs.

   A data migration has the following distinct moments:
\begin{enumerate}
   \item the moment of transition (the MoT), i.e. the moment that users experience changed functionality;
   \item the moment of completion (the MoC), i.e. the migration engineer removes the migration system and the old system, including its data.
\end{enumerate}

   Before the migration, the existing system satisfies its invariants (by definition).
   When the migration system compiles correctly, it satisfies its own invariants,
   so it can be deployed in production.
   That is the MoT.
   The task of the migration system is to satisfy the invariants of the desired system.
   Some of these invariants can be satisfied automatically by means of enforce rules.
   There may be invariants, however, that require human interaction to resolve.
   Once all invariants of the desired system are satisfied, the migration is ready for the MoC.
   So, the schema of the migration system must contain the rules to satisfy all invariants of the desired system.

   During the migration phase, transactions in the existing system are allowed because the migration engine will transport them to the desired system.
   However, after the MoT there must be no new changes in the existing system,
   to allow the system to finish transactions while the desired system is up and running.

@@

   \begin{definition}[Data set]
   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance \subseteq  {\Pair{\Atoms}{\Concepts}}$ that satisfies:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\instance A\ \wedge\ b\instance B
   \label{eqn:wellTypedEdge}
\end{eqnarray}
\end{definition}
   Looking at the example,
   equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that the Person Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.

@@

   For a different kind of constraint, the system is told what to do in order to fix it.
   Such constraints are called enforcement rules.
   We use just one type of enforcement rule\footnote{Ampersand has several types of enforcement rules.} in this work.
   Its syntax in Ampersand is as follows:
\begin{align}
      \text{\tt ENFORCE}\ r\ \text{\tt >:}\ \id{term}\label{enforce ins}
\end{align}

@@

   Let us look what happens at runtime.
   Initially, when {\tt copied\_isPartOf} is still empty, the difference between {\tt old.isPartOf} and {\tt copied\_isPartOf} equals {\tt old.isPartOf}.
   So, {\tt new.isPartOf} will contain the pairs of {\tt old.isPartOf} by rule~\ref*{difference}.
   Then, the engine inserts the same pairs into {\tt copied\_isPartOf} because of rule~\ref*{fill copies with new.isPartOf}.
   After this, triples with the relation {\tt new.isPartOf} can be inserted and deleted freely.
   Once all triples have been copied, {\tt copied\_isPartOf} will contain all pairs of {\tt old.isPartOf} and rule~\ref*{difference} will no longer insert any triples in {\tt new.isPartOf}.
   So, this rule can be removed after the copying has taken place.
   Consequently, declarations~\ref{extraRelation} thru~\ref{fill copies with new.isPartOf} can be removed from the migration script at the MoC.
